---
title: "medfatetraits - From plant trait data to model parameters"
output: github_document
---

```{r setup, echo = FALSE}

knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "README-")
```

  <!-- badges: start -->
  [![R-CMD-check](https://github.com/emf-creaf/medfatetraits/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/emf-creaf/medfatetraits/actions/workflows/R-CMD-check.yaml)
  <!-- badges: end -->

## Introduction

Package **medfatetraits** is designed to facilitate creating species parameter tables for package [medfate](https://github.com/emf-creaf/medfate). Specifically utility functions are provided to harmonize plant trait data bases and defining tables from them.

## Package installation

The latest stable versions GitHub as follows (required package `remotes`` should be installed/updated first):

```{r installation, eval = FALSE}
remotes::install_github("emf-creaf/medfatetraits")
```

## Usage

The package is still under active development. 

