---
title: "Allometry database harmonization"
author: "Miquel De Cáceres / Nicolas Martin-StPaul"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: TRUE
bibliography: traits4models.bib
params:
  spparams_rebuild: FALSE
  IFN_rebuild: FALSE
---

## Introduction

The availability of plant allometric equations continuously increases, as additional efforts are made in observational studies. This means that multiple data sources need to be harmonized before species parameter tables are build, in terms of:

    a) Nomenclature of the response and predictor variables.
    b) Allometric equation fitted.
    c) Taxonomy of the measured biological entities. 

The result of this harmonization needs to be stored in a harmonized format for subsequent use when creating species parameter tables. This vignette illustrates the harmonization procedures for an example data set using package **trait4models** and the usual **tidyverse** packages. Harmonization of plant trait data is explained in a companion vignette.

### Required packages

Assuming we have **traits4models** installed, we load it and other common packages that we will employ in this vignette:
```{r message=FALSE}
library(traits4models)
library(tidyverse)
library(readr)
```


### Example dataset

As an example of allometric equations, we will use a small compilation of crown width models, from different sources, that depend on diameter at breast height:

```{r, message=FALSE}
DB_path <- "~/OneDrive/EMF_datasets/AllometryDatabases/"
db <- openxlsx::read.xlsx(paste0(DB_path, "Sources/TreeAllometries/TreeAllometries.xlsx"),
                                 sheet= "Tree_CW_models")
```

The data looks as follows:
```{r}
db
```


## Harmonizing equations and notation

The allometric equation has two parameters that are used to estimate crown width from diameter at breast height (DBH). We start our harmonization by subsetting the necessary data:
```{r}
db_var <- db |>
  dplyr::rename(originalName = "Name",
                a = "a_cw",
                b = "b_cw",
                Reference = "Source") 
```

Importantly, for the allometry to be used, we need to specify the equation:
```{r}
db_var <- db_var |>
  dplyr::mutate(Equation = "CrownWidth = a·DBH^b")
```

We also add information on the response variable and predictor variable:
```{r}
db_var <- db_var |>
                dplyr::mutate(Response = "CrownWidth",
                ResponseDescription = "Crown width (m)",
                Predictor1 = "DBH",
                PredictorDescription1 = "Diameter at breast height (cm)")
```

If we want **traits4models** to correctly interpret the allometric equation at the stage of model parameter filling, it is important that columns `Equation` and `Response` are appropriately filled. Note, in addition, that the allometric equation starts with the name of the response on the left hand side of the equation.

Finally, we may add the priority of processing:
```{r}
db_var <- db_var |>
          dplyr::mutate(Priority = 1) |>
          dplyr::relocate(Reference, .after = PredictorDescription1)
```

The database after harmonizing allometry information looks as follows:
```{r}
db_var
```


## Taxonomic harmonization

Package **traits4models** currently relies on [World Flora Online](https://www.worldfloraonline.org/) for taxonomic harmonization, via package **WorldFlora** [@kindt_worldflora_2020] available at [CRAN](https://cran.r-project.org/package=WorldFlora). This latter package requires a static copy of the World Flora Online Taxonomic Backbone data that can be downloaded from the [World Flora Online](https://www.worldfloraonline.org/) website. Note that you should use a different DOI reference when reporting your harmonization procedures. We are using here [v.2024.06](https://doi.org/10.5281/zenodo.7460141). We assume the user has already downloaded the backbone and stored it in a file called *classification.csv*. 

```{r}
WFO_file <- paste0(DB_path, "../PlantTraitDatabases/WFO_Backbone/classification.csv")
```

Taxonomic harmonization is done by calling function `harmonize_taxonomy_WFO()` with a data frame (where notation and units are already harmonized) and the path to the WFO backbone (we omit the console output):

```{r, eval = FALSE}
db_post <- traits4models::harmonize_taxonomy_WFO(db_var, WFO_file)
```

```{r, message= FALSE, warning=FALSE, error=FALSE, echo = FALSE, include = FALSE}
db_post <- traits4models::harmonize_taxonomy_WFO(db_var, WFO_file)
```

The function requires that the input data frame contains a column called `originalName`, to identify the original taxa names (additional columns are simply transferred to the output). It performs both direct and fuzzy matching, which may lead to large processing time for large datasets. If we inspect the resulting data frame, we will see the additional columns, informing about accepted names and parent taxonomic entities:

```{r}
head(db_post)
```

## Checking harmonized allometry data

The packages includes function `check_harmonized_allometry()` to check whether a given data frame conforms in structure and content to what is later required for parameter table filling:

```{r}
check_harmonized_allometry(db_post)
```

In this case, the data set is ready to be used in parameter estimation. As an example, we can run the same checking function with the data base before taxonomic harmonization:

```{r}
check_harmonized_allometry(db_var)
```

## Storing harmonized dataset

Harmonized data tables can be stored in **.csv** text format or compressed **.rds** format. 

```{r, eval = FALSE}
write.csv2(db_post, "Products/harmonized/tree_crown_width_europe.csv", row.names = FALSE)
```
Note that all tables should be stored in the same directory, here in the `Products/harmonized/` path, for populate model parameters.

## Accessing harmonized allometry data

### List of harmonized allometry databases

As mentioned in the introduction, here it is assumed that a set of plant allometry databases have been harmonized. In our case, harmonized data files have been stored in the following path:

```{r}
harmonized_allometry_path = "~/OneDrive/EMF_datasets/AllometryDatabases/Products/harmonized"
```

We can list the set of harmonized sources using:
```{r}
allometry_files <- list.files(path = harmonized_allometry_path, full.names = FALSE)
allometry_files
```

### Querying allometry data for particular response variable

Before filling any species parameter table, it may be useful to inspect the amount of information available for particular response. Package **trait4models** provides function `get_allometry_data()`:

```{r}
fb_data <- get_allometry_data(harmonized_allometry_path, "FoliarBiomass",
                              progress = FALSE)
head(fb_data)
```


## References
