---
title: "Building species parameter tables for medfate"
author: "Miquel De Cáceres / Nicolas Martin-StPaul"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: TRUE
bibliography: traits4models.bib
params:
  spparams_rebuild: FALSE
  IFN_rebuild: FALSE
---

## Introduction

### Aim

This vignette describes the procedures used to create and fill species parameter data tables (such as `SpParamsES`, `SpParamsFR` and `SpParamsUS` included in package **traits4models**) for simulations with packages **medfate** and **medfateland**. Here it is assumed that a set of plant trait and allometry databases have been harmonized and will be used as source data for parameter estimation (see vignettes [Trait database harmonization](https://emf-creaf.github.io/traits4models/articles/TraitDatabaseHarmonization.html) and [Allometry database harmonization](https://emf-creaf.github.io/traits4models/articles/AllometryDatabaseHarmonization.html)). 

>**IMPORTANT**: This vignette is not self-contained, in the sense that it cannot be reproduced without access to data sets that are not included. Nevertheless, it is intended to serve as example of species parameterization for other regions.

### Required packages

Assuming we have **medfate** and **traits4models** correctly installed, we load them and other common packages that we will employ in this vignette:
```{r message=FALSE}
library(medfate)
library(traits4models)
library(tidyverse)
library(openxlsx)
library(sf)
```

### Target parameters

The set of species parameters needed to run the models included in **medfate** are described in a data frame called `SpParamsDefinition` included in the package. It is obvious that **medfate** requires a sheer number of parameters cannot be searched for manually, even for a small number of target species. Hence, it is important to draw values from plant allometry and plant trait data bases. Not all model parameters have their counterpart in publicly-available databases, and even when using data bases it is unlikely that one will find appropriate parameter values for all species. To help with this situation, **medfate** has *inbuilt imputation procedures* to estimate missing values for many parameters before starting model simulations. However, imputed parameter estimates will normally be worse than estimates coming from data bases. Hence, process-based model users should put a strong effort in finding source parameter data before starting simulations.

### Steps to build a species parameter table

The following sections describe different steps that we used to obtain the species parameter table, which can be grouped into six main steps:

| # | Step                                                  | Function(s)         |
|---|-------------------------------------------------------|---------------------|
| 1 | Initialize the parameter table with target taxonomic entities | `init_medfate_params()` |
| 2 | Populate species parameters from forest inventory data | `fill_medfate_inventory_traits()` |
| 3 | Populate plant allometric coefficients from suitable databases |  `fill_medfate_allometries()`|
| 4 | Populate plant functional traits from harmonized data bases |  `fill_medfate_traits()`|
| 5 | Checking the final parameter table | `check_medfate_params()` |
| 6 | If necessary, completing strict parameters | `complete_medfate_strict()` |

The remainder of this vignette illustrates each of these steps.

## Initializing a species parameter table

### Target taxonomic entities

The species parameter table for **medfate** starts with a vector of species names. Here, we will assume that those names correspond to taxa (genus, species, subspecies, ...) mentioned in a national forest inventory, more specifically the Spanish National Forest Inventory (SFI). Note that taxon names in a forest inventory will correspond with concepts of their own taxonomic reference. Hence, it is important to harmonize taxonomy with the reference used in **traits4models**, i.e. [World Flora Online](https://www.worldfloraonline.org/), for which function `harmonize_taxonomy_WFO()` may be useful. 

Here we will assume that such harmonization has been conducted for the SFI taxon list that is included as external data in the package:

```{r}
file <- system.file("extdata", "NFI_ES_mapping.csv", package = "traits4models")
```

```{r}
NFI_ES_mapping <- read.table(file, sep=";", header=TRUE, na.strings = "") |>
  tibble::as_tibble() |>
  dplyr::mutate(NFICode = as.character(NFICode))
NFI_ES_mapping
```

Here `NFIName` corresponds to the name used in the Spanish forest inventory, while `originalName` contains the names that where used for taxonomy harmonization with function `harmonize_taxonomy_WFO()`, and `acceptedName` corresponds to the accepted name according to World Flora Online. Column `originalName`  has the same values as `NFIName` for a scientific taxa, but is otherwise missing. Non-scientific plant names (e.g. 'Cultivated trees', 'Other conifers') or taxa names that could not be identified cannot be used in medfate simulations because they cannot be matched with plant trait data. Normally, we will need to filter those names using:

```{r}
spp_filt <- NFI_ES_mapping |>
  dplyr::filter(!is.na(acceptedName)) 
```

### Initialization

Package **traits4models** provides function `init_medfate_params()` to initialize a species parameter table for **medfate** according to the names given in a data frame with taxonomic information, here `NFI_ES_mapping`. Column `originalName` will be used to match the `Species` names in `forest` objects derived from the Spanish forest inventory, while column `acceptedName` will be used to ensure that matching with trait information is done using the appropriate (harmonized) taxonomic nomenclature. Other columns are used to fill in taxonomy:

```{r, eval = TRUE}
SpParams <- traits4models::init_medfate_params(spp_filt,
                                               verbose = FALSE)
```

The result is an empty data frame, except for the first columns, which correspond to taxonomic hierarchy of each entity (genus, family, group, ...), where taxonomic information from lower rank (up to family) comes from the input data frame; and that of higher rank (order, group) is completed using internal package data obtained from package **taxize**. The number of rows in the output data frame can be larger than the input, because by default the function creates additional rows for species (when only sub-species of that species have been cited) or genus (when species of that genus have been cited).

```{r}
SpParams |> tibble::as_tibble()
```


## Filling structural parameters from forest inventory data

Some species parameters can be withdrawn from the data of the target forest inventory where simulations are to be conducted. We load a data frame (actually an `sf` object) containing data from the Third Spanish Forest Inventory:

```{r}
sf_IFN3 <- readRDS("~/OneDrive/mcaceres_work/model_initialisation/medfate_initialisation/IFN2medfate/data/SpParamsES/IFN3/soilmod/IFN3_spain_soilmod_WGS84.rds")
sf_IFN3
```
Column `forest` contains the forest inventory plots that will be used to withdraw parameters. For example, the second `forest` object is:

```{r}
sf_IFN3$forest[[2]]
```

**traits4models** provides function `fill_medfate_inventory_traits()` to perform the parameter extraction:

```{r}
SpParams<- traits4models::fill_medfate_inventory_traits(SpParams, sf_IFN3, 
                                                        progress = FALSE)
```

Species matching is performed between column `Species` of the tree or shrub data tables and `Name` of the species parameter table; and the following information is extracted:

* `GrowthForm`: Growth form according to the usage in the forest inventory. For example, if the species is cited in `treeData` tables but not in `shrubData` tables, then growth form will be `"Tree"`.
* `Hmax`: Maximum tree/shrub height (cm), according to the `Height` column in `treeData` or `shrubData` and `quantile_Hmax` parameter.
* `Hmed`: Median tree/shrub height (cm), according to the `Height` column in `treeData` or `shrubData` and `quantile_Hmed` parameter.
* `fHDmin`, `fHDmax`: Minimum or maximum height to diameter ratio for trees, according to the `Height` and `DBH` columns in `treeData` and `quantile_fHDmin` or `quantile_fHDmax` parameters, respectively.

## Filling allometric coefficients

As mentioned in the introduction, here it is assumed that a set of plant allometry databases have been harmonized. In our case, harmonized data files have been stored as **.rds** or **.csv** files in the following path:

```{r}
harmonized_allometry_path = "~/OneDrive/EMF_datasets/AllometryDatabases/Products/harmonized"
```

Function `fill_medfate_allometries()` performs trait parameter filling for **medfate** using the available data:

```{r}
SpParams<- traits4models::fill_medfate_allometries(SpParams, 
                                                   harmonized_allometry_path, 
                                                   progress = TRUE, verbose = FALSE)
```

## Filling parameters from harmonized trait databases

As mentioned in the introduction, here it is assumed that a set of plant trait databases have been harmonized. In our case, harmonized data files have been stored as **.rds** or **.csv** files in the following path:

```{r}
harmonized_trait_path <- "~/OneDrive/mcaceres_work/model_development/medfate_parameterization/Traits&Models/data/harmonized_trait_sources"
```

Function `fill_medfate_traits()` performs trait parameter filling for **medfate** using the available data:

```{r}
SpParams<- traits4models::fill_medfate_traits(SpParams, harmonized_trait_path, 
                                              progress = FALSE, verbose = FALSE)
```

Parameter filling sometimes implies translating trait data into the units required for model parameters. The function identifies the row of the species parameter table to modify by matching column `AcceptedName` of `SpParams` with the column `acceptedName` of the harmonized trait data sources. When multiple values of a given species are available, the function takes the median value (for quantitative traits) or the mode (i.e. the most frequent value; for qualitative traits). In addition, data sources of higher priority are processed first. Thus, data sources of lower priority being used for those cases not covered by prioritized sources.

## Checking species parameter tables

Before using a given species parameter table, it is important to verify that the table does not contain missing values in key parameters. In medfate, those parameters that are always required non-missing values are called `Strict` (see `SpParamsDefinition` in package **medfate**). Function `check_medfate_params()` checks, among other things, that strict parameters do not contain missing values:

```{r}
check_medfate_params(SpParams)
```

Normally, the information of plant trait data bases will not contain information for all species and strict parameters, even if most of them are qualitative traits. This problem is raised by `check_medfate_params()` and must be addressed.

## Completing strict parameters

Package **trait4models** includes function `complete_medfate_strict()` to perform imputation of strict parameters, so that the result table can be used for simulations. Imputation is performed on the basis of available information for similar taxa. Taxonomically closer entities are given priority over higher order taxa (i.e. species of the same genus are priorized as sources before searching for species in the same family):

```{r}
SpParams <- traits4models::complete_medfate_strict(SpParams, progress = FALSE)
```

After completion of strict paratemers, the species parameter table should pass the previous check:
```{r}
check_medfate_params(SpParams)
```

And we can store the table for its use in model simulations.

## References
