---
title: "Trait database harmonization"
author: "Miquel De Cáceres / Nicolas Martin-StPaul"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: TRUE
bibliography: traits4models.bib
params:
  spparams_rebuild: FALSE
  IFN_rebuild: FALSE
---

## Introduction

Species parameter values cannot be drawn from a single data source. Moreover, the availability of plant trait data continuously increases, as additional efforts are made in observational or experimental studies [@Kattge2020]. This means that multiple data sources need to be harmonized before species parameter tables are build, in terms of:

    a) Nomenclature of measured variables.
    b) Measurement units.
    c) Taxonomy of the measured biological entities. 

The result of this harmonization needs to be stored in a harmonized format for subsequent use when creating species parameter tables. This vignette illustrates the harmonization procedures for an example data set using package **trait4models** and the usual **tidyverse** packages. Harmonization of allometric equations is explained in a companion vignette.

>**IMPORTANT**: This vignette is not self-contained, in the sense that it cannot be reproduced without access to data sets that are not included. Nevertheless, it is intended to serve as example of species parameterization for other regions.

### Required packages

Assuming we have **traits4models** installed, we load it and other common packages that we will employ in this vignette:
```{r message=FALSE}
library(traits4models)
library(tidyverse)
library(readr)
```


### Example dataset

As example for the harmonization process, here we will use data from a Bartlett et al. [-@Bartlett2016]. Much larger data sets can (and should) be processed, but they take much more time. Bartlett et al. provide traits that describe the leaf/stem/root hydraulic vulnerability curves of several species as well as water potential corresponding to stomatal aperture. We start by loading the dataset:

```{r, message=FALSE}
DB_path <- "~/OneDrive/EMF_datasets/PlantTraitDatabases/"
db <- readr::read_csv(paste0(DB_path, "Sources/Bartlett_et_al_2016/pnas.1604088113.sd01.csv"))
```

The data looks as follows:
```{r}
db
```


## Harmonizing notation and measurement units

The first steps to be done are to harmonize trait notation, i.e. how plant traits are referred to, and if necessary, change their units. Package **traits4models** includes a data table called `HarmonizedTraitDefinition` that presents plant trait definitions and their required notation and units:


```{r, echo = FALSE}
data("HarmonizedTraitDefinition")
knitr::kable(HarmonizedTraitDefinition)
```

In the case of Bartlett's dataset we are interested in points of the vulnerability curve and stomatal behavior. We can check their corresponding names in `HarmonizedTraitDefinition`. For example, `Leaf P50 (MPa)` is the water potential corresponding to the 50% conductance loss in leaves, and should be named `VCleaf_P50` according to `HarmonizedTraitDefinition`, and so on. We can use **dplyr** function `rename()` to harmonize trait notation. In this case, all plant traits we are interested in are given in units of MPa, and `HarmonizedTraitDefinition` reports the same units for these traits, so there is no need to harmonize measurement units. The code for notation harmonization could be as follows:

```{r}
db_var <- db |>
  dplyr::select(Name, "Leaf P50 (MPa)", "Stem P50 (MPa)", "Stem P88 (MPa)", "Stem P12 (MPa)", 
                "Root P50 (MPa)", "Gs P50 (MPa)", "Gs 95 (MPa)") |>
  dplyr::rename(originalName = Name,
                VCleaf_P50 = "Leaf P50 (MPa)",
                VCstem_P50 = "Stem P50 (MPa)",
                VCstem_P12 = "Stem P12 (MPa)",
                VCstem_P88 = "Stem P88 (MPa)",
                VCroot_P50 = "Root P50 (MPa)",
                Gs_P50 = "Gs P50 (MPa)",
                Gs_P95 = "Gs 95 (MPa)") 
```

Note that we also renamed the column containing the plant species into `originalName`. The original name represents the taxon name that is used by the data set owner/provider and is key for taxonomic harmonization. The result of this step should contain `originalName`, plus one column for each harmonized trait, and (preferably) three columns called `Reference`, `DOI` and `Priority`. We can add those columns manually using:

```{r}
db_var <- db_var |>
  dplyr::mutate(Reference = "Bartlett et al. (2016). The correlations and sequence of plant stomatal, hydraulic, and wilting responses to drought. PNAS 113: 13098-13103",
                DOI = "10.1073/pnas.1604088113",
                Priority = 3)
```

Columns `Reference` and `DOI` indicates the bibliographic source of the data, whereas `Priority` allows defining an order in which trait data sources will be processed. Those with highest priority order (lowest value of `Priority`) will be given preference.

```{r}
db_var
```


## Taxonomic harmonization

Package **traits4models** currently relies on [World Flora Online](https://www.worldfloraonline.org/) for taxonomic harmonization, via package **WorldFlora** [@kindt_worldflora_2020] available at [CRAN](https://cran.r-project.org/package=WorldFlora). This latter package requires a static copy of the World Flora Online Taxonomic Backbone data that can be downloaded from the [World Flora Online](https://www.worldfloraonline.org/) website. Note that you should use a different DOI reference when reporting your harmonization procedures. We are using here [v.2024.06](https://doi.org/10.5281/zenodo.7460141). We assume the user has already downloaded the backbone and stored it in a file called *classification.csv*. 

```{r}
WFO_file <- paste0(DB_path, "WFO_Backbone/classification.csv")
```

Taxonomic harmonization is done by calling function `harmonize_taxonomy_WFO()` with a data frame (where notation and units are already harmonized) and the path to the WFO backbone (we omit the console output):

```{r, eval = FALSE}
db_post <- traits4models::harmonize_taxonomy_WFO(db_var, WFO_file)
```

```{r, message= FALSE, warning=FALSE, error=FALSE, echo = FALSE, include = FALSE}
db_post <- traits4models::harmonize_taxonomy_WFO(db_var, WFO_file)
```

The function requires that the input data frame contains a column called `originalName`, to identify the original taxa names (additional columns are simply transferred to the output). It performs both direct and fuzzy matching, which may lead to large processing time for large datasets. If we inspect the resulting data frame, we will see the additional columns, informing about accepted names and parent taxonomic entities:

```{r}
head(db_post)
```


## Checking harmonized trait data

The packages includes function `check_harmonized_trait()` to check whether a given data frame conforms in structure and content to what is later required for parameter table filling:

```{r}
check_harmonized_trait(db_post)
```

In this case, the data set is ready to be used in parameter estimation. As an example, we can run the same checking function with the data base before taxonomic harmonization:

```{r}
check_harmonized_trait(db_var)
```

## Storing harmonized dataset

Harmonized data tables can be stored in **.csv** text format or compressed **.rds** format. Moreover, all tables should be stored in the same directory, here in the `Products/harmonized/` path.

```{r, eval = FALSE}
file_out <- paste0(DB_path, "Products/harmonized/Bartlett_et_al_2016.rds")
saveRDS(db_post, file_out)
```

## Accessing harmonized trait data

### List of harmonized trait databases

As mentioned in the introduction, here it is assumed that a set of plant trait databases have been harmonized. In our case, harmonized data files have been stored in the following path:

```{r}
harmonized_trait_path <- "~/OneDrive/EMF_datasets/PlantTraitDatabases/Products/harmonized"
```

We can list the set of harmonized sources using:
```{r}
trait_files <- list.files(path = harmonized_trait_path, full.names = FALSE)
trait_files
```

### Querying data for particular traits or species

Before filling any species parameter table, it may be useful to inspect the amount of information available for particular traits or species. Package **trait4models** provides a couple of utility functions for this. For example, we can load all values for Gswmin, the minimum stomatal conductance, using function `get_trait_data()`:

```{r}
gsmin_data <- get_trait_data(harmonized_trait_path, "Gswmin",
                           progress = FALSE)
head(gsmin_data)
```

Analogously, if we are interested in querying trait information for a particular taxa, we can use function `get_taxon_data()`:

```{r}
ph_data <- get_taxon_data(harmonized_trait_path, "Pinus halepensis",
                          progress = FALSE)
head(ph_data)
```

## References
