---
title: "Trait database harmonization"
author: "Miquel De Cáceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: TRUE
bibliography: traits4models.bib
params:
  spparams_rebuild: FALSE
  IFN_rebuild: FALSE
---

## Introduction

Species parameter values cannot be drawn from a single data source. This means that multiple data sources need to be harmonized before species parameter tables are build, in terms of:

    a) Nomenclature of measured variables.
    b) Measurement units.
    c) Taxonomy of the measured biological entities. 

The result of this harmonization needs to be stored in a harmonized format for subsequent use when creating species parameter tables. This vignette illustrates the harmonization procedures for an example data set.

>**IMPORTANT**: The vignette is not self-contained, in the sense that it cannot be reproduced without access to data sets that are not included. Nevertheless, it is intended to serve as example of species parameterization for other regions.

### Required packages

Assuming we have **medfate** and **traits4models** installed, we load them and other common packages that we will employ in this vignette:
```{r message=FALSE}
library(medfate)
library(traits4models)
library(tidyverse)
library(readr)
```

### Harmonized trait definition, notation and units

The set of species parameters needed to run the models included in **medfate** are described in a dataframe called `HarmonizedTraitDefinition`: 



```{r, echo = FALSE}
data("HarmonizedTraitDefinition")
knitr::kable(HarmonizedTraitDefinition)
```

It is obvious that such a sheer number of parameters cannot be searched for manually, even for a small number of target species. Hence, it is important to draw values from allometric and plant trait data bases. Not all model parameters have their counterpart in publicly-available databases, and even when using data bases it is unlikely that one will find appropriate parameter values for all species. To help with this situation, **medfate** has *inbuilt imputation procedures* to estimate missing values for many parameters before starting model simulations. However, imputed parameter estimates will normally be worse than estimates coming from data bases. Hence, we should put a strong effort in finding source parameter data before relying on inbuilt imputation.


### Example dataset

```{r, message=FALSE}
DB_path <- "~/OneDrive/EMF_datasets/PlantTraitDatabases/"
db <- readr::read_csv(paste0(DB_path, "Sources/Bartlett_et_al_2016/pnas.1604088113.sd01.csv"))
```

```{r}
db
```

 
## Harmonizing notation and measurement units



```{r}
db_var <- db |>
  dplyr::select(Name, "Leaf P50 (MPa)", "Stem P50 (MPa)", "Stem P88 (MPa)", "Stem P12 (MPa)", 
                "Root P50 (MPa)", "Gs P50 (MPa)", "Gs 95 (MPa)") |>
  dplyr::rename(originalName = Name,
                VCleaf_P50 = "Leaf P50 (MPa)",
                VCstem_P50 = "Stem P50 (MPa)",
                VCstem_P12 = "Stem P12 (MPa)",
                VCstem_P88 = "Stem P88 (MPa)",
                VCroot_P50 = "Root P50 (MPa)",
                Gs_P50 = "Gs P50 (MPa)",
                Gs_P95 = "Gs 95 (MPa)") 
```


## Taxonomic harmonization

```{r}
WFO_file <- paste0(DB_path, "WFO_Backbone/classification.csv")
db_post <- traits4models::harmonize_taxonomy_WFO(db_var, WFO_file)
```


## Checking and storing harmonized dataset

The packages includes function `check_harmonized_table()` to check whether a given data frame conforms in structure and content to what is later needed for parameter table filling:

```{r}
check_harmonized_table(db_post)
```

In this case, columns `Units` and `Reference` are missing, but they are not enforced and the data set is ready to be used in parameter estimation.

Harmonized data tables should be preferably stored in compressed **.rds** format. Moreover, all tables should be stored in the same directory, here in the `Products/harmonized/` path.

```{r, eval = FALSE}
file_out <- paste0(DB_path, "Products/harmonized/Bartlett_et_al_2016.rds")
saveRDS(db_post, file_out)
```


## References
