---
title: "Trait database harmonization"
author: "Miquel De Cáceres"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: TRUE
bibliography: medfatetraits.bib
params:
  spparams_rebuild: FALSE
  IFN_rebuild: FALSE
---

## Introduction

Species parameter values cannot be drawn from a single data source. This means that multiple data sources need to be harmonized before species parameter tables are build, in terms of:

    a) Nomenclature of measured variables.
    b) Measurement units.
    c) Taxonomy of the measured biological entities. 

The result of this harmonization needs to be stored in a harmonized format for subsequent use when creating species parameter tables. This vignette illustrates the harmonization procedures for an example data set.

>**IMPORTANT**: The vignette is not self-contained, in the sense that it cannot be reproduced without access to data sets that are not included. Nevertheless, it is intended to serve as example of species parameterization for other regions.

### Required packages

Installed version of **medfate** should be **ver 2.9** or higher. 
```{r, eval = FALSE}
install.packages("medfate")
```

If the version at CRAN is older, the package should be installed from GitHub via `remotes::install_github()`:
```{r, eval = FALSE}
remotes::install_github("emf-creaf/medfate")
```

Package **medfatetraits** is required for the species parameterization process, as it includes several functions that facilitate parsing data sources into parameter values. Package **medfatetraits** is installed only from GitHub:

```{r, eval = FALSE}
remotes::install_github("emf-creaf/medfatetraits")
```

Once we have **medfate** and **medfatetraits**, we load them and other common packages that we will employ in this vignette:
```{r message=FALSE}
library(medfate)
library(medfatetraits)
# library(IFNread)
library(tidyverse)
library(openxlsx)
```

### Target parameters

The set of species parameters needed to run the models included in **medfate** are described in a dataframe called `SpParamsDefinition`: 


```{r}
data("SpParamsDefinition")
SpParamsDefinition$ParameterName
```
It is obvious that such a sheer number of parameters cannot be searched for manually, even for a small number of target species. Hence, it is important to draw values from allometric and plant trait data bases. Not all model parameters have their counterpart in publicly-available databases, and even when using data bases it is unlikely that one will find appropriate parameter values for all species. To help with this situation, **medfate** has *inbuilt imputation procedures* to estimate missing values for many parameters before starting model simulations. However, imputed parameter estimates will normally be worse than estimates coming from data bases. Hence, we should put a strong effort in finding source parameter data before relying on inbuilt imputation.



 
## Harmonizing measurement units

## Taxonomic harmonization

## Storing harmonized databases

## References
